---
title: "Quarto HTML Basics"
author: "Veronica Vargas"
date: "`r format(Sys.time(), '%e %B %Y')`"
toc: true
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    theme: minty
execute:
  warning: false
---

## Install

1.  Download and install the latest release of **R** and **RStudio** (Note that these are two different things and you should install both):

    ::: {.callout appearance="minimal"}
    <i class="bi bi-download"></i> [Download R and RStudio](https://posit.co/download/rstudio-desktop/)
    :::

    ::: callout-warning
    ## Note

    If you have already installed R and RStudio from previous classes, make sure your versions are up to date.
    :::

2.  Next, download and install the latest release of **Quarto**, we will use this for creating the submission document:

    ::: {.callout appearance="minimal"}
    <i class="bi bi-download"></i> [Download Quarto](https://quarto.org/docs/get-started/)
    :::

3.  Read through the Quarto tutorials at [Get Started](https://quarto.org/docs/get-started/hello/rstudio.html). You can also use **VSCode** if that is the IDE you are familiar with.

4.  Finally, we will install a few packages necessary for this class. We will use `tidyverse` for plotting, `car` and `lmtest` for regression diagnostics, and `ISLR`, `MASS`, and `AER` for datasets. Open Rstudio, and use `install.packages()` function in R to install them:

    ``` r
    install.packages("tidyverse")
    install.packages("car")
    install.packages("lmtest")
    install.packages("ISLR")
    install.packages("MASS")
    install.packages("AER") 
    ```

5.  Now, download the `lab0.qmd` file from Canvas and open it.

## Text and Math

Quarto is based on Pandoc and uses its variation of markdown as its underlying document syntax. Pandoc markdown is an extended and slightly revised version of John Gruber's [Markdown](https://daringfireball.net/projects/markdown/) syntax.

Markdown is a plain text format that is designed to be easy to write, and, even more importantly, easy to read:

> A Markdown-formatted document should be publishable as-is, as plain text, without looking like it's been marked up with tags or formatting instructions. -- [John Gruber](https://daringfireball.net/projects/markdown/syntax#philosophy)

Below are some examples of formatting with Pandoc Markdown. See the full documentation of [Quarto's Guide](https://quarto.org/docs/authoring/markdown-basics.html) for more in-depth documentation.

### Text Formatting

+-----------------------------------------+-----------------------------------------+
| Markdown Syntax                         | Output                                  |
+=========================================+=========================================+
| ``` markdown                            | *italics*, **bold**, ***bold italics*** |
| *italics*, **bold**, ***bold italics*** |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` markdown                            | superscript^2^ / subscript~2~           |
| superscript^2^ / subscript~2~           |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` markdown                            | ~~strikethrough~~                       |
| ~~strikethrough~~                       |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` markdown                            | `verbatim code`                         |
| `verbatim code`                         |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` markdown                            | -   unordered list                      |
| * unordered list                        |     -   sub-item 1                      |
|     + sub-item 1                        |     -   sub-item 2                      |
|     + sub-item 2                        |         -   sub-sub-item 1              |
|         - sub-sub-item 1                |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` markdown                            | 1.  ordered list                        |
| 1. ordered list                         | 2.  item 2                              |
| 2. item 2                               |     i)  sub-item 1                      |
|     i) sub-item 1                       |         A.  sub-sub-item 1              |
|          A.  sub-sub-item 1             |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+

### Tables

We can manually create tables in Quarto with Markdown as follow:

#### Markdown Syntax

``` markdown
| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|   12  |  12  |    12   |    12  |
|  123  |  123 |   123   |   123  |
|    1  |    1 |     1   |     1  |
```

#### Output

| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|    12 | 12   | 12      |   12   |
|   123 | 123  | 123     |  123   |
|     1 | 1    | 1       |   1    |

Learn more in the article on [Tables](tables.qmd).

### Math

Quarto uses $\LaTeX$ ([Latex Project](https://www.latex-project.org/)) to render mathematical equations. This is the standard way of writing math in academia and elsewhere. You can either display the math *inline*, i.e. within normal text, using a pair of dollar signs `$`. For example, this code:

```         
The data consists of $n$ observations $\{x_i, y_i\}^n_{y=1}$
```

Will show up as:

"The data consists of $n$ observations $\{x_i, y_i\}^n_{y=1}$."

Or, you can have *display equations*, which are on separate lines, using double dollar signs `$$`:

```         
$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i
$$
```

Will show up as:

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i
$$ Preferably, you will use $\LaTeX$ to write the equations in answers to your non-programming questions and submit everything in same file. Below are some example latex scripts and their outputs. Refer to [this cheatsheet](https://tug.ctan.org/info/undergradmath/undergradmath.pdf) for more examples.

+-----------------------------------------+-----------------------------------------+
| $\LaTeX$ Syntax                         | Output                                  |
+=========================================+=========================================+
| ``` latex                               | $$                                      |
| $$                                      | \frac{1}{2}                             |
| \frac{1}{2}                             | $$                                      |
| $$                                      |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$                                      |
| $$                                      | Y = X_1 + X_2                           |
| Y = X_1 + X_2                           | $$                                      |
| $$                                      |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$                                      |
| $$                                      | a^2 + b^2 = c^2                         |
| a^2 + b^2 = c^2                         | $$                                      |
| $$                                      |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$                                      |
| $$                                      | \mathbf{(X^{\prime}X)^{-1}X^{\prime}y}  |
| \mathbf{(X^{\prime}X)^{-1}X^{\prime}y}  | $$                                      |
| $$                                      |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$                                      |
| $$                                      | Y \sim X\beta_0 + X\beta_1 + \epsilon\\ |
| Y \sim X\beta_0 + X\beta_1 + \epsilon\\ | \epsilon \sim N(0,\sigma^2)             |
| \epsilon \sim N(0,\sigma^2)             |  $$                                     |
| $$                                      |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$x =  \begin{cases}                    |
| $$x = \begin{cases}                     | 1 \;\;\text{ if } \mu > 0\\             |
| 1 \;\;\text{ if } \mu > 0\\             | 0 \;\;\text{ otherwise }                |
| 0 \;\;\text{ otherwise }                | \end{cases}$$                           |
| \end{cases}$$                           |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+
| ``` latex                               | $$\begin{bmatrix}                       |
| $$\begin{bmatrix}                       | 1 & 2 & 3\\                             |
| 1 & 2 & 3\\                             | a & b & c                               |
| a & b & c                               | \end{bmatrix}$$                         |
| \end{bmatrix}$$                         |                                         |
| ```                                     |                                         |
+-----------------------------------------+-----------------------------------------+

## Computation

In this section, we will look at basic computation and results presentation in R and Quarto. Let's use the `Credit` dataset from the **ISLR** package for our example.

```{r}
#| label: load-packages
library(ISLR)
library(ggplot2)
```

First, we can look at the data.

```{r}
head(Credit)
```

We can use the `knitr::kable()` function to print tables as HTML, for better looking tables:

```{r}
knitr::kable(head(Credit))
```

Then, we can summarize the dataset:

```{r}
summary(Credit)
```

As we can see, the output are printed out once we render the document.

Now, let's visualize the relationship between Credit rating (`Rating`), and Income in 1000s (`Income`), by gender (`Gender`), using `ggplot()` and `geom_point()`:

```{r}
#| label: scatterplot

ggplot(Credit, aes(x = Rating, y = Income, col = Gender)) +
  geom_point(alpha = 0.5, size = 2) +
  theme_minimal()
```

If we just want to embed the plot (or any other types of output), we can add `#| code-fold: true` to the code block to hide the code:

```{r}
#| label: scatterplot_folded
#| code-fold: true

ggplot(Credit, aes(x = Rating, y = Income, col = Gender)) +
  geom_point(alpha = 0.5, size = 2) +
  theme_minimal()
```

Quarto can also display interactive plots. For example, we can use `plotly` package to make this interactive:

```{r}
#| label: scatterplot_ly

library(plotly)
p <- ggplot(Credit, aes(x = Rating, y = Income, col = Gender)) +
  geom_point(alpha = 0.5, size = 2) +
  theme_minimal()
ggplotly(p)
```

Now, to get a taste of what you will learn in the class, let's go ahead and estimate a regression model of Credit rating (`Rating`) on income (`Income`) and gender (`Gender`), and display the results, with `lm()`. Don't worry about interpreting the output right now!

```{r}
ols <- lm(Rating ~ Income  + Gender, Credit)
summary(ols)
```

The above is a typical output of a liner regression model. You will see a lot of these this semester.

## Your Turn

Now, to make sure you have properly installed R, RStudio, and Quarto, let's answer a few quick questions:

1.  Create an R code block below this line, and type in `version`.

    ```{r}
    `version`
    ```

2.  Write down the formula for variance of a random variable ([See here](https://en.wikipedia.org/wiki/Variance) if you forget) in $\LaTeX$ below this line.

    $$
    Var[X] = E[(X-\mu)^2]
    $$

3.  Use `ggplot` to create a scatterplot of credit limit (`Limit`), age (`Age`), and colored by ethnicity (`Ethnicity`). Embed the plot below this line, fold the code.

    ```{r}
    #| code-fold: true

    ggplot(Credit, aes(x = Age, y = Limit, col = Ethnicity)) +
      geom_point(alpha = 0.5, size = 2) +
      theme_minimal()
    ```

4.  Once you are done with the above, change `Author:` field at the beginning of the document to your name, click `Render` button at the top, and submit the HTML file to Canvas.

::: callout-note
## Note

The HTML file will be in the same folder as the `lab0.qmd` file.
:::
